|

### 2021-12-16 06-Chp-IPD-Analysis.Rmd
Proofreading edits

Previous work has suggested a difference in the effect of lipids on dementia risk based on participant age and sex.[@ancelin2013; @mielke2010] However, the systematic review presented in Chapters \@ref(sys-rev-methods-heading) & \@ref(sys-rev-results-heading) did not identify strong evidence for an effect of sex or age at lipid measurement on the relationship between lipids and dementia. The number of included studies in the lipid fraction meta-analyses was small, due to both the relatively small number of studies reporting on the lipids exposure and the poor reporting of summary statistics of participant characteristics in those that did. Additionally, best practice guidance recommends against basing the decision to perform an IPD meta-analysis on the between-study heterogeneity in a summary data meta-analysis, as similar distributions of participant covariates across studies may mask a true effect.[@riley2020]

As such, the aims of this analysis are two fold: firstly, to perform an IPD meta-analysis across identified cohorts to examine the impact of participant age-at-measurement and sex on the relationship between lipids and dementia risk; and secondly, to expand the evidence base for the effect of lipids on dementia outcomes by obtaining estimates from previously unanalysed cohorts available via the Dementia Platform UK, a large consortium of dementia cohorts.

@@

I considered four blood lipid fractions as part of this analysis, namely: total cholesterol (TC), low-density lipoprotein cholesterol (LDL-c), high-density lipoprotein cholesterol (HDL-c) and triglycerides (TG). Cohorts were eligible for inclusion if they contained data on at least one of these recorded as a continuour variable (i.e. studies with binary "hypercholesteromemia" would be excluded).

@@

Potentially eligible data sources as defined above were identified via two mechanisms, each with a distinct focus. For both approaches, the number of cohorts responding to the request for data access and, where applicable, the reasons given for a refusal were recorded.

@@

The first approach focused on previously analysed observational prospective cohort studies examining the effect of blood lipid levels on dementia outcomes, as identified by the systematic review presented in Chapters \@ref(sys-rev-methods-heading) & \@ref(sys-rev-results-heading). The data sources used in each of these cohort analyses were screened against the criteria listed in the previous section, and eligible cohorts were approached for data access. In the first instance, the first/corresponding author of the publication was emailed <!-- TODO (see Appendix \@ref(ipd-email-collab) for a copy of the correspondence sent to potential collaborators) --> in Autumn 2020. If this approach did not elicit a response within two months, the last author was contacted, on the basis that the first/corresponding author may have been a more junior member of the research group who had moved to a different institution.

@@

The second approach focused on incorporating relevant, previously-unanalysed data into the analysis, thus providing additional evidence on the relationship between blood lipids and dementia risk. This was achieved through the Dementia Platform UK (DPUK), a collaborative grouping of existing dementia cohorts established by the MRC which works with data owners to make their data readily accessible for secondary analysis.[@bauermeister2020] It provides access to 42 cohorts with over 3 million participants, and makes use of a central streamlined application process for all cohorts with the intent of making it easier to access data from existing data sources.

@@

Where data on one of the exposures (TC, LDL-c, HDL-c, TG) was missing, it was inferred from the other three fractions (where available) using the Friedwald formula (Equation \@ref(eq:total-cholesterol-formula)). Lipids levels were retained as continuous variables rather than dichotomising into a binary hypercholesterolemia exposure, given the additional statistical power this adds to the meta-analysis.[@riley2020;@ensor2018] Where lipid measurements were reported in _mg/dL_, these were converted to _mmol/L_.

@@

A range of additional variables were included in the analysis, intended to address the potential for confounding. With an awareness that discrepancies in available covariates is a common issue in IPD studies, I defined an idealised set of covariate domains to be age, sex, education, BMI, _Apo_$\mathcal{E}4$ status, smoking/alcohol status, ethnicity, and prevalent diabetes or cardiovascular disease. This set covers key risk factors for dementia/Alzheimer's disease in addition to general cardiovascular risk factors. Details on how these variables were coded, given the available data, are presented in Section \@ref(ipd-covar-definition).

@@

Missing data in this analysis was classified as either relating to missing values (a data on a variable of interest was available in a cohort, but some values were missing) or missing variables (a cohort did not collect data on a variable of interest). Variables with missing values were identified, and 20 imputed datasets were created.[@sterne2009] Imputation was performed using the MICE (Multiple Imputation by Chained Equations) using the `mice` package in R.

@@

<!-- TODO May need to rethink this, as introducing model based on missing data which isn't fully discussed until later. -->

@@

In terms of the analytic approach taken, a two-stage IPD analysis was used. Under a two-stage approach, estimates for the effect of each lipid fraction on incident dementia were first calculated for each data source. More specifically, a logistic regression model adjusted for relevant covariates as detailed in the above section was used to quantify the effect of a 1 _mmol/L_ increase in each lipid fraction on each dementia outcome.

@@

<!-- REVIEW ORs - the one issue here may be the length of follow-up. Was it roughly equal across cohorts such that the ORs are comparable? -->

A two-stage IPD approach was employed for a number of reasons. Firstly, and most importantly, a two-stage approach allows for siloed data, enabling researchers who are unwilling/unable to provide their data to btain the effect estimates themselves, following a specified analysis plan, and share these with the IPD team.[@riley2010] Secondly, it allows for the production of forest plots of within-cohort estimates, something which is useful for the triangulation exercise in Chapter \@ref(tri-heading). Finally, a two-stage approach is simpler to model, as it uses standard well-documented summary effect estimate meta-analysis techniques, and automatically accounts for methodological issues such as clustering within cohorts[@abo-zaid2013] and the potential for ecological bias.[@burke2017]

@@

In order to investigate the interaction of participant-level characteristics (age and sex) with lipid levels, lipid-covariate interaction terms were extracted and synthesised using a fixed effects meta-analysis.[@fisher2017] All interaction analysis were performed using the common-set-adjusted model (Model 1). To avoid ecological bias, where the between-study association does not reflect the within-study associations, cohorts where there was no within-study variation in the covariate of interest were excluded from the interaction analysis for that covariate.[@burke2017] As an example, it is impossible to estimate the impact of sex on the lipid/dementia relationship in a study which contains only female participants.

@@

Of the `r n_applied$total` studies to which I applied for data access, only three (`r paste0(comma((n_accessed$total/n_applied$total)*100),"%")`) were included in the final analysis. Figure \@ref(fig:cohortFlowchart) details whether the cohorts eventually included in the review were identified by the systematic review or via the DPUK portal. In addition, the reasons for cohorts not being included in the analysis are presented, where available, stratified by application approach.

In summary, the requests for data from cohorts identified by the systematic review were characterised by a very low response rate (N = 5, 25%). For the minority who did respond, common reasons given by authors for not sharing the data included that they: no longer worked with the same group and did not know if the data was available or how to obtain it; no longer had access to the data; or were currently performing, or intended to perform, a similar analysis as the one proposed.

@@

(ref:cohortFlowchart-cap) __Flowchart detailing sources of data for the IPD analysis__ The number of cohorts approached and the reasons for exclusion are summarised, stratified by identification method (systematic review vs DPUK).

@@

Nine cohorts (8 DPUK only, 1 Systematic review + DPUK) responded positively to requests for data access. However, on inspection of data provided, six of these cohorts were excluded and the reason for exlcusion in each case is summarised in Table \@ref(tab:dataExcluded-table). In summary, two cohorts contained were memory-complaint cohorts, containing participants unlikely to be free of dementia at baseline (BRACE, MEMENTO). Two cohorts, despite being designed as multi-wave studies, only shared data related to a single wave and so represented cross-sectional data (Generation Scotland, NICOLA). Finally TRACK HD is a genetic cohort where cohort owners advised me that dementia will inevitably develop in mid-life driven by the Huntington's (HTT gene) mutation and that this effect would likely far outweigh the effect of lipids. Finally, the ELSA study was excluded based on the dichotomous definition of the exposure variable. <!-- TODO CITATION NEEDED for each study here! -->

@@

The Caerphilly Prospective Study (CaPS) is a longitudinal study of men in South Wales, UK.[@zotero-15398;@elwood2013] Blood lipids (TC, LDL-c, HDL-c and TG) were measured at baseline in 1979-1983, and from Phase III (1989-1993) onwards, a battery of cognitive tests were introduced. A dementia was derived using data obtained during Phase V (2002-2004), giving between 19-25 years of follow-up. The cohort contains data on dementia outcomes sub-classified as vascular and non-vascular dementia. Data was available for all covariates of interest except for ethnicity, BMI, prevalent IHD and _Apo_$\mathcal{E}4$ status. While the CaPS study collected data on height/weight (from which BMI can be calculated as $\frac{weight}{height^2}$) and prevalent heart disease, these were not available from the DPUK version of the CaPS data.

@@

The European Prospective Investigation of Cancer (EPIC) - Norfolk is a population-based cohort, containing men and women recruited from 35 general practices in Norfolk between 1993 and 1998.[@riboli1997; @riboli2002] Data provided by this cohort for the analysis, dementia was ascertained at at the 5th Health Checkup (2016-2018), providing between 18-25 years of follow-up. The added evidential value of the EPIC cohort is small, given the fact that the data obtained contains only 8 dementia events. The cohort contains no information on dementia subtype, while all covariates of interest bar _Apo_$\mathcal{E}4$ were available.

@@

The Whitehall II study is a prospective cohort study of men and women recruited between 1985 and 1989 from the civil service in London.[@marmot2005] Lipids measurements were available from the third wave of the study, conducted between 1991-1993. The cohort is linked with the Hospital Episode Statistics (HES) database, a database containing details of participant events at NHS hospitals in England, which was used to define covariates at baseline and dementia outcomes.[@zotero-15403] HES data for dementia outcomes was available up to March 2015, providing between 22-24 years of follow-up. Information was available on dementia type, which was classified as all-cause dementia, Alzheimer's disease or vascular dementia. This data source contained details on all covariates of interest except for _Apo_$\mathcal{E}4$ status.

@@

Based on available data across cohorts, smoking was classified as never/ever/current, while alcohol consumption was classified as never/ever. Education was categorised into 4 levels (None, O-levels, A-levels, Degree). BMI was treated as a continuous variable while presence of vascular co-morbidities was treated as dichotomous.

@@

Missing values in collected variables was common across the cohorts. A matrix of covariates, and whether they were not available or contained missing values for each included cohort, can be seen Appendix \@ref(appendix-ipd-covariate-matrix). The Whitehall II and EPIC cohorts contained data on all but one covariates of interest, while three missing covariates were identified in the CaPS cohort, namely education, prevalent ischemic heart disease, BMI and ethnicity. No included cohort provided information on the _Apo_$\mathcal{E}4$ status of participants, and so this variable could not be adjusted for in the analysis.

@@

Across the three cohorts, `r n_total_ipd` participants were included in this analysis (Whitehall II = 8208, EPIC = 1115, CaPS = 2512). All cohorts contained data on the four exposure variables of interest (or sufficient data from which to calculate them) and on all-cause dementia outcomes. The only other dementia outcome examined across cohorts was vascular dementia, which was reported in the CaPS and Whitehall II studies. The definitions of dementia outcomes used across cohorts can be seen in Appendix \@ref(appendix-ipd-dementia-def). Cumulatively, there were `r n_ipd_dementia` cases of all-cause dementia, with `r n_ipd_vasdem` further classified as vascular dementia. Summary statistics for each cohort are provided in Table \@ref(tab:covariateSummary-table).

@@

<!----------------------------------------------------------------------------->
(ref:covariateSummary-caption) __Characteristics of IPD cohorts__ - Summary of characteristics for cohorts included in the IPD analysis. Variables not available from a cohort are denoted by "-". See Appendix \@ref(appendix-ipd-covariate-matrix) for details on the level of missing data within collected variables.

@@

The results from the main effect analysis across the varying lipid fractions on each dementia outcome considered can be seen in Figures \@ref(fig:mainEffectDem) & \@ref(fig:mainEffectVad), respectively. There was weak evidence for an association of any lipid level with either all-cause or vascular dementia, with the exception of a harmful association between raised triglycerides and vascular dementia (`r ipd_vasdem`). For the sole cohort containing data on the Alzheimer's disease outcome (Whitehall II), there was weak evidence for an association of this outcome with any lipid fraction (results shown in comparison with previous analysis of this cohort in Figure \@ref(fig:whitehallComparisonAd))

@@

The maximally-adjusted model (Model 2) could have been employed in the analysis of the effect of sex, as the CaPS cohort contains only men and had to be excluded from this analysis, leaving the EPIC and Whitehall cohorts, both of which had a full complement of covariates. However, I decided to employ the same underlying model across the interaction analyses to aid comparability between the age and sex interaction estimates.

For all-cause dementia, there was no evidence of an interaction of lipid levels with either age group or sex (Figures \@ref(fig:interactionDementiaAge) & \@ref(fig:interactionDementiaSex)). Note that for the sex interaction analysis, the CaPS cohort was excluded as it contains only a single sex and therefore provides no information on exposure-sex interaction.

<!----------------------------------------------------------------------------->
(ref:interactionDementiaAge-cap)  __Meta-analysis of age-lipid interaction terms for all-cause dementia__ - Age was grouped into 5-year age bands, and the effect estimatse presented represent the OR per 1-step increase in age group. Estimates were obtained using the common-set-adjusted model (Model 1: age, sex, smoking, alcohol, education, diabetes).

@@

<!----------------------------------------------------------------------------->
(ref:interactionDementiaSex-cap) __Meta-analysis of sex-lipid interaction terms for all-cause dementia__ - Estimates were obtained using the common-set-adjusted model (Model 1: age, sex, smoking, alcohol, education, diabetes) and refer to the effect of male gender on the lipid/all-cause dementia association.

@@

<!----------------------------------------------------------------------------->
(ref:interactionVascularAge-cap) __Meta-analysis of age-lipid interaction terms for vascular dementia__ - Age was grouped into 5-year age bands, and the effect estimatse presented represent the OR per 1-step increase in age group. Estimates were obtained using the common-set-adjusted model (Model 1: age, sex, smoking, alcohol, education, diabetes).

@@

This analysis requested data from `r n_applied$total` data sources, but only obtained data from three, all of which were based in the United Kingdom. No evidence for an effect of lipids on the risk of dementia or related outcomes was identified, except for a harmful association of raised triglycerides with risk of vascular dementia. Similarly, there was weak evidence for an interaction of the effect of lipid levels on dementia outcomes with participants' age, grouped into 5 year bands, or sex.

@@

The obvious key limitation of this analysis is the very low response rate to requests for data access, which may bias the results if there are systematic differences in the association of lipids with dementia outcomes between cohorts that share data and those that do not.[@ahmed2012] Whether or not to press ahead with an IPD analysis in the absence of all (or even most) data is a personal decision, and some previous analyses have highlighted where they decided not to pursue an IPD analysis.[@jaspers2014] For the purposes of this thesis, the decision was made to conduct the IPD analysis, as it provided training in application of the methods in addition to providing new evidence that will be incorporated into the triangulation exercise detailed in Chapter \@ref(tri-heading).

<!-- REVIEW also identified that any future studies would require a longer and/or more directive approach to the provision of access - DPUK is a good example of how this could be done! -->

@@

The obstacles to data access described above are in theory what the DPUK was built to address. However, even with the help of the streamlined application process afforded by the DPUK, accessing sufficient data was a challenge. The response rate among DPUK cohorts a year after application was just 50%. In addition, some cohorts responded saying that the proposed study question was already under investigation by another group, and that they would not share the data on this basis. In light of this, a centralised database of ongoing analyses being performed using DPUK data would be of enormous help, reducing research waste and providing opportunities for collaboration. Finally, the DPUK process would be aided by a clearer distinction between those cohorts that are "DPUK native" (i.e. where a copy of the data is already held on DPUK servers) versus externally hosted, as the response time for externally hosted cohorts is likely to be much longer.

@@

More generally, systematically missing variables required a trade-off between inclusion of cohort data and appropriate control for confounding. The final choice of the common-set-adjusted model for use in the interaction analyses mean there is the potential for residual confounding. However, sensitivity analysis comparing the common-set-adjusted (Model 1) and fully-adjusted (Model 2) models in cohorts with a full complement of covariates indicated that further adjustment for BMI, ethnicity, education and prevalent IHD had minimal impact on the effect estimates. Of note, this is comparable with the analysis of the CPRD data presented in the previous chapter, where adjustment for variable beyond age and sex had a limited impact on the observed effect estimates (see Section \@ref(cprd-impact-additional-covar)).

@@

Investigation revealed several possible reasons for this discrepancy. The first is that while this analysis uses multiple imputation to address missing data, the Tynkkynen _et al._ analysis does not describe how missing data was handled. However, the reduced precision in the reported estimates suggest a complete case analysis. This interpretation is supported by a comparison of summary statistics, which illustrated that my analysis had substantially more dementia events (N =287 in this analysis vs. n=114 in Tynkkynen _et al._), suggesting that incomplete records were discarded in the previous analysis. Finally, the covariates adjusted for in each analysis were similar, though the previous analysis had access to genetic data allowing it to adjust for _Apo_$\mathcal{E}4$ status. However, given that _Apo_$\mathcal{E}4$ is a risk factor for increased LDL-c rather than triglycerides,[@bennet2007] it seems unlikely that additional adjustment for this variable is responsible for the discrepancy in findings observed.

@@

While this analysis did not manage to obtain and analyse a large proportion of identified data, a central strength of this analysis is the use of a systematic approach to identify and attempt to contact relevant cohorts. Further, it also enabled incorporation of two previously unanalysed datasets - the CaPS and EPIC Norfolk cohorts - thus providing additional evidence that is used in the triangulation exercise reported in Chapter \@ref(tri-heading).

@@

Finally, this analysis provides new evidence on a previously unexplored outcome (vascular dementia), adding to the extremely small evidence base for this outcome identified by the systematic review presented in Chapters \@ref(sys-rev-methods-heading) & \@ref(sys-rev-results-heading). This is relevant even in the previously analysed WhiteHall II cohort, as the previous analysis by Tynkkynen _et al._ did not report on this outcome.

@@

Additionally, the production of detailed guidance for handling cases where covariates are systematically missing would represent a useful addition to the topic. Much of the literature around IPD analysis is focused on the synthesis of RCTs, where additional covariate information is needed primarily for the assessment of treatment-covariate interactions rather than the adjustment of the effect estimate for confounding. Given the wide availability of non-randomised cohorts, improved guidance on this challenge would support future work.

Finally, a movement towards increased use of unique persistent identification of researchers, through schemes such as the ORCID programme,[@nature2009] would help with contact issues. Researchers move institutions regularly as their contracts come to an end, and so the institutional contact details provided on publications are frequently out of date.

>>>>

Previous work has suggested a difference in the effect of lipids on dementia risk based on participant age and sex.[@ancelin2013; @mielke2010] The systematic review presented in Chapters \@ref(sys-rev-methods-heading) & \@ref(sys-rev-results-heading) could not investigate this effect because the number of included studies in the lipid fraction meta-analyses was small, due to both the relatively small number of studies reporting on lipids and the poor reporting of summary statistics of participant characteristics in those that did. Additionally, best practice guidance recommends against basing the decision to perform an IPD meta-analysis on between-study heterogeneity in a summary data meta-analysis, because similar distributions of participant covariates across studies may mask a true effect.[@riley2020]

As such, the aims of this analysis are two fold. Firstly, I plan to perform an IPD meta-analysis across identified cohorts to examine the impact of participant age-at-measurement and sex on the relationship between lipids and dementia risk. Secondly, I aim to expand the evidence base for the effect of lipids on dementia outcomes by obtaining estimates from previously unanalysed cohorts available via the Dementia Platform UK, a large consortium of dementia cohorts.

@@

I considered four blood lipid fractions as part of this analysis, namely: total cholesterol (TC), low-density lipoprotein cholesterol (LDL-c), high-density lipoprotein cholesterol (HDL-c) and triglycerides (TG). Cohorts were eligible for inclusion if they contained data on at least one of these lipid fractions,  recorded as a continuous variable (i.e. studies with binary "hypercholesteromemia" would be excluded).

@@

Potentially eligible data sources were identified via two approaches, each with a distinct focus. The approaches are described in detail in the following sections. For both approaches, the number of cohorts responding to the request for data access and, where applicable, the reasons given for a refusal were recorded.

@@

The first approach focused on previously analysed observational prospective cohort studies examining the effect of blood lipid levels on dementia outcomes, as identified by the systematic review presented in Chapters \@ref(sys-rev-methods-heading) & \@ref(sys-rev-results-heading). The data sources used in each of these cohort analyses were screened against the criteria listed in the previous section, and eligible cohorts were approached for data access. In the first instance, the first/corresponding author of the publication was emailed in Autumn 2020. If this approach did not elicit a response within two months, the last author was contacted, on the basis that the first/corresponding author may have been a more junior member of the research group who had moved to a different institution.

@@

The second approach focused on incorporating relevant, previously-unanalysed data into the analysis, thus providing additional evidence on the relationship between blood lipids and dementia risk. This was achieved through the Dementia Platform UK (DPUK), a collaborative grouping of existing dementia cohorts established by the Medical Research Council which works with data owners to make their data readily accessible for secondary analysis.[@bauermeister2020] It provides access to 42 cohorts with over 3 million participants, and makes use of a central streamlined application process for all cohorts, with the intent of making it easier to access data from existing data sources.

@@

Where data on one of the exposures (TC, LDL-c, HDL-c, TG) was missing, it was inferred from the other three fractions (where available) using the Friedwald formula (Equation \@ref(eq:total-cholesterol-formula)). Lipid levels were retained as continuous variables rather than dichotomising into a binary hypercholesterolemia exposure, given the additional statistical power this adds to the meta-analysis.[@riley2020;@ensor2018] Where lipid measurements were reported in _mg/dL_, these were converted to _mmol/L_.

@@

A range of additional variables were included in the analysis, intended to address the potential for confounding. With an awareness that discrepancies are common in the set of available covariates across cohorts included in an IPD analysis, I defined an idealised set of covariate domains to be age, sex, education, BMI, _Apo_$\mathcal{E}4$ status, smoking/alcohol status, ethnicity, and prevalent diabetes or cardiovascular disease. This set covers key risk factors for dementia/Alzheimer's disease in addition to general cardiovascular risk factors. Details on how these variables were coded, given the available data, are presented in Section \@ref(ipd-covar-definition).

@@

Missing data in this analysis was classified as either relating to missing values (a data on a variable of interest was available in a cohort, but some values were missing) or missing variables (a cohort did not collect data on a variable of interest). Variables with missing values were identified, and 20 imputed datasets were created.[@sterne2009] Imputation was performed using MICE (Multiple Imputation by Chained Equations) and was implemented in R  using the `mice` package in R.[@Van_Buuren2011-nc]

@@

@@

In terms of the analytic approach taken, a two-stage IPD analysis was used. Under a two-stage approach, estimates for the effect of each lipid fraction on incident dementia were first calculated for each data source. More specifically, a logistic regression model adjusted for relevant covariates (as detailed in the above section) was used to quantify the effect of a 1 _mmol/L_ increase in each lipid fraction on each dementia outcome.

@@

A two-stage IPD approach was employed for a number of reasons. Firstly, and most importantly, a two-stage approach allows for siloed data, enabling researchers who are unwilling/unable to provide their data to obtain the effect estimates themselves, following a specified analysis plan, and share these with the IPD team.[@riley2010] Secondly, it allows for the production of forest plots of within-cohort estimates, something which is useful for the triangulation exercise in Chapter \@ref(tri-heading). Finally, a two-stage approach is simpler to model because it uses standard well-documented summary effect estimate meta-analysis techniques, and automatically accounts for methodological issues such as clustering within cohorts[@abo-zaid2013] and the potential for ecological bias.[@burke2017]

@@

In order to investigate the interaction of participant-level characteristics (age and sex) with lipid levels, lipid-covariate interaction terms were extracted and synthesised using a fixed effects meta-analysis.[@fisher2017] All interaction analyses were performed using the common-set-adjusted model (Model 1). To avoid ecological bias, where the between-study association does not reflect the within-study associations, cohorts where there was no within-study variation in the covariate of interest were excluded from the interaction analysis for that covariate.[@burke2017] As an example, it is impossible to estimate the impact of sex on the lipid/dementia relationship in a study which contains only female participants.

@@

Of the `r n_applied$total` studies to which I applied for data access, only three (`r paste0(comma((n_accessed$total/n_applied$total)*100),"%")`) were included in the final analysis. Figure \@ref(fig:cohortFlowchart) details whether the cohorts eventually included in the review were identified by the systematic review or via the DPUK portal. In addition, the reasons for cohorts being from the analysis are presented, stratified by application approach.

In summary, the requests for data from cohorts identified by the systematic review were characterised by a very low response rate (N = 5, 25%). For the five cohorts that did respond, common reasons given by authors for not sharing the data included that they: no longer worked with the same group and did not know if the data was available or how to obtain it; no longer had access to the data; or were currently performing, or intended to perform, a similar analysis as the one proposed.

@@

(ref:cohortFlowchart-cap) __Flowchart detailing sources of data for the IPD analysis__ The number of cohorts approached and the reasons for exclusion are stratified by identification method (systematic review vs DPUK).

@@

Nine cohorts (8 DPUK only, 1 Systematic review + DPUK) responded positively to requests for data access. However, on inspection of data provided, six of these cohorts were excluded and the reason for exclusion in each case is shown in Table \@ref(tab:dataExcluded-table). In summary, two cohorts contained were memory-complaint cohorts, containing participants unlikely to be free of dementia at baseline (BRACE, MEMENTO). Two cohorts, despite being designed as multi-wave studies, only shared data related to a single wave and so represented cross-sectional data (Generation Scotland, NICOLA). Finally TRACK HD is a genetic cohort where cohort owners advised me that dementia will inevitably develop in mid-life driven by the Huntington's (HTT gene) mutation and that this effect would likely far outweigh the effect of lipids. Finally, the ELSA study was excluded based on the dichotomous definition of the exposure variable. <!-- TODO CITATION NEEDED for each study here! -->

@@

The Caerphilly Prospective Study (CaPS) is a longitudinal study of men in South Wales, UK.[@zotero-15398;@elwood2013] Blood lipids (TC, LDL-c, HDL-c and TG) were measured at baseline in 1979-1983, and from Phase III (1989-1993) onwards, a battery of cognitive tests were introduced. Dementia outcomes were determined using data obtained during Phase V (2002-2004), giving between 19-25 years of follow-up. The cohort contains data on dementia outcomes sub-classified as vascular and non-vascular dementia. Data was available for all covariates of interest except for ethnicity, BMI, prevalent IHD and _Apo_$\mathcal{E}4$ status. While the CaPS study collected data on height/weight (from which BMI can be calculated as $\frac{weight}{height^2}$) and prevalent heart disease, these were not available from the DPUK version of the CaPS data.

@@

The European Prospective Investigation of Cancer (EPIC) - Norfolk is a population-based cohort, containing men and women recruited from 35 general practices in Norfolk between 1993 and 1998.[@riboli1997; @riboli2002] Dementia was ascertained at the 5th Health Check-up (2016-2018), providing between 18-25 years of follow-up. The added evidential value of the EPIC cohort is small, given the fact that the data obtained contains only 8 dementia events. The cohort contains no information on dementia subtype, while all covariates of interest bar _Apo_$\mathcal{E}4$ were available.

@@

The Whitehall II study is a prospective cohort study of men and women recruited between 1985 and 1989 from the civil service in London.[@marmot2005] Lipids measurements were available from the third wave of the study, conducted between 1991-1993. The cohort is linked with the Hospital Episode Statistics (HES) database, a database containing details of participant events at NHS hospitals in England. which was used to capture dementia outcomes.[@zotero-15403] HES data was available up to March 2015, providing between 22-24 years of follow-up. Information was available on dementia type, which was classified as all-cause dementia, Alzheimer's disease or vascular dementia. This data source contained details on all covariates of interest except for _Apo_$\mathcal{E}4$ status.

@@

Based on available data across cohorts, smoking was classified as never/ever/current, while alcohol consumption was classified as never/ever. Ever use in this case refers to participants who do not currently smoke/drink alcohol, but did so in the past. Education was categorised into 4 levels (None, O-levels, A-levels, Degree). BMI was treated as a continuous variable while presence of vascular co-morbidities was treated as dichotomous.

@@

Missing values in collected variables was common across the cohorts. A matrix of covariates for each included cohort, describing both missing variables and the proportion of missing values within collected variables, can be seen in Appendix \@ref(appendix-ipd-covariate-matrix). The Whitehall II and EPIC cohorts contained data on all but one covariates of interest, while three missing covariates were identified in the CaPS cohort, namely education, prevalent ischemic heart disease, BMI and ethnicity. No included cohort provided information on the _Apo_$\mathcal{E}4$ status of participants, and so this variable could not be adjusted for in the analysis.

@@

Across the three cohorts, `r n_total_ipd` participants were included in this analysis (Whitehall II = 8208, EPIC = 1115, CaPS = 2512). All cohorts contained data on the four lipid fractions of interest (or sufficient data from which to calculate them) and on all-cause dementia outcomes. The only other dementia outcome examined across cohorts was vascular dementia, which was reported in the CaPS and Whitehall II studies. The definitions of dementia outcomes used across cohorts can be seen in Appendix \@ref(appendix-ipd-dementia-def). Cumulatively, there were `r n_ipd_dementia` cases of all-cause dementia, with `r n_ipd_vasdem` further classified as vascular dementia. Summary statistics for each cohort are provided in Table \@ref(tab:covariateSummary-table).

@@

<!----------------------------------------------------------------------------->
(ref:covariateSummary-caption) __Characteristics of IPD cohorts__ - Summary of characteristics for cohorts included in the IPD analysis. Variables not available from a cohort are denoted by "-". See Appendix \@ref(appendix-ipd-covariate-matrix) for details on the proportion of missing data within collected variables.

@@

The results from the main effect analysis across the varying lipid fractions on each dementia outcome considered can be seen in Figures \@ref(fig:mainEffectDem) & \@ref(fig:mainEffectVad), respectively. There was weak evidence for an association of any lipid level with either all-cause or vascular dementia, with the exception of a harmful association between raised triglycerides and vascular dementia (`r ipd_vasdem`, \@ref(fig:mainEffectVad)). For the sole cohort containing data on the Alzheimer's disease outcome (Whitehall II), there was weak evidence for an association of this outcome with any lipid fraction (results shown in comparison with previous analysis of this cohort in Figure \@ref(fig:whitehallComparisonAd))

@@

The maximally-adjusted model (Model 2) could have been employed in the analysis of the effect of sex, as both EPIC and Whitehall II had a full complement of covariates (CaPS contains only men and so was excluded from this analysis). However, I decided to employ the same underlying model across the interaction analyses to aid comparability between the age and sex interaction estimates.

For all-cause dementia, there was no evidence of an interaction of lipid levels with either age group or sex (Figures \@ref(fig:interactionDementiaAge) & \@ref(fig:interactionDementiaSex)).

<!----------------------------------------------------------------------------->
(ref:interactionDementiaAge-cap)  __Meta-analysis of age-lipid interaction terms for all-cause dementia__ - Age was grouped into 5-year age bands, and the effect estimates presented represent the OR per 1-step increase in age group. Estimates were obtained using the common-set-adjusted model (Model 1: age, sex, smoking, alcohol, education, diabetes).

@@

<!----------------------------------------------------------------------------->
(ref:interactionDementiaSex-cap) __Meta-analysis of sex-lipid interaction terms for all-cause dementia__ - Estimates were obtained using the common-set-adjusted model (Model 1: age, sex, smoking, alcohol, education, diabetes) and refer to the effect of male gender on the lipid/all-cause dementia association. The CaPS cohort was excluded from this analysis because it contains only a single sex and therefore provides no information on the lipid-sex interaction.

@@

<!----------------------------------------------------------------------------->
(ref:interactionVascularAge-cap) __Meta-analysis of age-lipid interaction terms for vascular dementia__ - Age was grouped into 5-year age bands, and the effect estimates presented represent the OR per 1-step increase in age group. Estimates were obtained using the common-set-adjusted model (Model 1: age, sex, smoking, alcohol, education, diabetes).

@@

This analysis requested data from `r n_applied$total` data sources, but only obtained data from three, all of which were based in the United Kingdom. No evidence for an effect of lipids on the risk of dementia or related outcomes was identified, except for a harmful association of raised triglycerides with risk of vascular dementia. Similarly, there was weak evidence for an interaction of the effect of lipid levels on dementia outcomes with participants' age (grouped into 5 year bands) or sex.

@@

The obvious key limitation of this analysis is the very low response rate to requests for data access, which may bias the results if there are systematic differences in the association of lipids with dementia outcomes between cohorts that share data and those that do not.[@ahmed2012] Whether or not to press ahead with an IPD analysis in the absence of all (or even most) data is a personal decision, and some previous analyses have highlighted where they decided not to pursue an IPD analysis.[@jaspers2014] For the purposes of this thesis, the decision was made to conduct the IPD analysis because it provided training in application of IPD methods in addition to providing new evidence that will be incorporated into the triangulation exercise detailed in Chapter \@ref(tri-heading).

@@

The obstacles to data access described above are in theory what the DPUK was built to address. However, even with the help of the streamlined application process afforded by the DPUK, accessing sufficient data was a challenge. The response rate among DPUK cohorts a year after application was just 50%. In addition, some cohorts responded saying that the proposed study question was already under investigation by another group, and that they would not share the data on this basis. In light of this, a centralised database of ongoing analyses being performed using DPUK data would be of enormous help, reducing research waste and providing opportunities for collaboration. Finally, the DPUK process would be aided by a clearer distinction between those cohorts that are "DPUK native" (i.e. where a copy of the data is already held on DPUK servers) versus externally hosted, seeing as the response time for externally hosted cohorts is likely to be much longer.

@@

More generally, systematically missing variables required a trade-off between inclusion of cohort data and appropriate control for confounding. The final choice of the common-set-adjusted model for use in the interaction analyses means there is the potential for residual confounding. However, sensitivity analysis comparing the common-set-adjusted (Model 1) and fully-adjusted (Model 2) models in cohorts with a full complement of covariates indicated that further adjustment for BMI, ethnicity, education and prevalent IHD had minimal impact on the effect estimates. Of note, this is comparable with the analysis of the CPRD data presented in the previous chapter, where adjustment for variables beyond age and sex had a limited impact on the observed effect estimates (see Section \@ref(cprd-impact-additional-covar)).

@@

Investigation revealed several possible reasons for this discrepancy. The first is that while this analysis uses multiple imputation to address missing data, the Tynkkynen _et al._ analysis does not describe how missing data was handled. However, the reduced precision in the reported estimates suggest a complete case analysis. This interpretation is supported by a comparison of summary statistics, which illustrated that my analysis had substantially more dementia events (N =287 in this analysis vs. n=114 in Tynkkynen _et al._), suggesting that incomplete records were discarded in the previous analysis. Finally, the covariates adjusted for in each analysis were similar, though the previous analysis had access to genetic data allowing it to adjust for _Apo_$\mathcal{E}4$ status. However, given that _Apo_$\mathcal{E}4$ is a risk factor for increased LDL-c rather than triglycerides,[@bennet2007] it seems unlikely that additional adjustment for this variable is responsible for the discrepancy in the findings observed.

@@

While this analysis did not manage to obtain and analyse a large proportion of identified data, a central strength of this analysis is the use of a systematic approach to identify and attempt to contact relevant cohorts. Furthermore, it also enabled incorporation of two previously unanalysed datasets - the CaPS and EPIC Norfolk cohorts - thus providing additional evidence that is used in the triangulation exercise reported in Chapter \@ref(tri-heading).

@@

Finally, this analysis provides new evidence on a previously unexplored outcome (vascular dementia), adding to the extremely small evidence base for this outcome identified by the systematic review presented in Chapters \@ref(sys-rev-methods-heading) & \@ref(sys-rev-results-heading). This is relevant even in the previously analysed WhiteHall II cohort because the previous analysis by Tynkkynen _et al._ did not report on this outcome.

@@

Additionally, the production of detailed guidance for handling cases where covariates are systematically missing would represent a useful contribution to the topic. Much of the literature around IPD analysis is focused on the synthesis of RCTs, where additional covariate information is needed primarily for the assessment of treatment-covariate interactions rather than the adjustment of the effect estimate for confounding. Given the wide availability of non-randomised cohorts, improved guidance on this challenge would support future work.

Finally, movement towards increased use of unique and persistent researcher identifiers, such as those offered by the ORCID programme,[@nature2009] would help with contact issues. Researchers move institutions regularly as their contracts come to an end, and so the institutional contact details provided on publications are frequently out of date.
